A Mountaintop View Requires Minimal Sorting: \break An Instance Optimal Contour Tree Algorithm

### 2014-10-24 paint-contour-tree.tex
proof in Section 7

\newcommand{\build}{{\tt build}}
\newcommand{\col}{col}
\newcommand{\fcol}{full}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\cost}{\mathop{cost}}
\newcommand{\h}{att}
\newcommand{\lift}{{\tt lift}}
\newcommand{\mcol}{mcol}
\newcommand{\merge}{{\tt merge}}
\newcommand{\onestep}{{\tt piece}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\pmax}{P_{\max}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{\cC}
\newcommand{\redH}{\widetilde{H}}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

We use a key theorem from~\cite{Carr} (Theorem 6.6) that maps paths in $\cC(\MM)$ and $\MM$.

\begin{theorem} \label{thm:carr-path} For every path $P$ in $\MM$, there exists a path $Q$ in the contour tree to which $P$
maps, and for every path $Q$ in the contour tree, there exists at least one path $P$ in $\MM$ that maps to $Q$.
\end{theorem}

@@

{\bf Painting:} The central tool is a notion 
of \emph{painting} of $\MM$. Initially associate a color with each maximum. Denoting the set of maxima is $X$,
this is an arbitrary bijection $\chi:X \mapsto [|X|]$. Imagine there being a large
can of paint of color $\chi(x)$ at maxima $x$. We will spill different paint from each maximum and watch it flow down.
This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
\emph{So paint only flows down edges, and it does not color the interior of facets.} Furthermore, paints
do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color. This process (and indeed the entire algorithm)
works purely on the 1-skeleton of $\MM$, which is just a graph.
%
\begin{definition} \label{def:paint} Let the 1-skeleton of $\MM$ be denoted $G(\MM) = (\cV,E)$.
A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following properties. 
\begin{asparaitem}
	\item The restriction $\chi:X \mapsto [|X|]$ is a bijection.
	\item Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
	is well-defined to be the unique maximum with the same color as $e$.)
\end{asparaitem}
\end{definition}
%
Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(|\MM|)$ time, using a descending BFS from each maximum that does not explore previously colored edges.

@@

\begin{definition} \label{def:color-full} Consider vertex $v$, and let $\phi$ be a $(f(v)+\delta)$-contour in
$f|_{B_\eps(v)}$. If $\phi$ is monochromatic of color $c$, then $v$ is \emph{fully touched by $c$}. The set of colors
that fully touch $v$ is denoted $\fcol(v)$.
\end{definition}

Note that $\fcol(v)$ has size at most $2$, by the Morse condition. Note that $\col(v)$ and $\fcol(v)$ can
be initially determined in $O(|\MM|)$ time, by a simple linear scan.

\subsection{The data structures} \label{sec:struct}

%
%We will store each set $\touch(c)$ in a binomial heap, keyed by the heights. 
Abusing notation, $\touch(c)$ refers
both to the set and the data structure used to store it.

\medskip
\noindent
{\bf The binomial heaps $\touch(c)$:} For each color $c$, $T(c) = \{v | c \in \fcol(v)\}$.
This is stored as a \emph{binomial max-heap} $T(c)$, keyed by vertex heights.

@@

\medskip
\noindent
{\bf Attachment vertex $\h(c)$:} For each color $c$, we maintain a critical point $\h(c)$ of this color.
We will maintain the guarantee that the portion of the contour tree above (and including) $\h(c)$ has already been constructed.

\subsection{Properties of a painted simplicial complex} \label{sec:prop}

\begin{lemma} \label{lem:full} Consider vertex $v$ and take $(f(v)+\delta)$-contour $\phi$ intersecting
$B_\eps(v)$. Suppose the edges incident to $v$ intersecting $\phi$ do not have the same color. For any such
color $c$, there exists a critical vertex $w$ strictly higher than $v$ such that $c \in \fcol(v)$.
\end{lemma}

\begin{proof} Consider the highest
vertex $w$ that is touched by $c$ and some other color. Clearly such a vertex exists, since the condition
is also satisfied by $v$. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
and intersect some $c$-colored edge incident to $w$. Note that $\phi$ is also colored $c$,
since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
edges from the maximum to $w$. For any point on this path, the contour passing through this point must
be colored $c$.) Hence, $c \in \fcol(w)$ and $w \neq v$. Furthermore, $w$ is a critical point,
since $w$ is touched by another color. The corresponding edge cannot intersect $\phi$, so $w$
must have up-degree $2$.
\end{proof}

\begin{definition} \label{def:ripe} A vertex $v$ is \emph{ripe} if: for all $c \in \fcol(v)$, $v$
is the highest vertex in $T(c)$. Also, for all $c \in \col(v) \setminus \fcol(v)$, $T(c)$ is empty.
\end{definition}

@@

\smallskip
\begin{compactenum}
	\item If $K$ is empty, push arbitrary unprocessed critical point $v$.
	\item Let $h$ be head of $\st$.
	\item While $h$ is not ripe:
	\begin{compactenum}
		\item (Either (i) there is $c \in \fcol(v)$ such that $v$ is not the highest in $T(c)$ or (ii) there is 
		$c \in \col(v) \setminus \fcol(v)$, such that $T(c)$ is non-empty.)
		\item In either case, push the highest of $T(c)$ onto $\st$, and update head $h$.
	\end{compactenum}
\end{compactenum}
\end{minipage}}

\medskip
Clearly, when $\update(\st)$ terminates, the head $h$ is ripe. The following crucial claim demonstrates
the utility of the stack.

@@

\begin{proof} Consider a single push onto the stack. In Case (i), the increase in height is obvious.
Consider Case (ii). There is a vertex $v$ touched by color $c \notin \fcol(v)$. Consider the highest
vertex $w$ that is touched by $c$ and some other color. Clearly such a vertex exists, since the condition
is also satisfied by $v$. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
and intersect some $c$-colored edge incident to $w$. Note that $\phi$ is also colored $c$,
since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
edges from the maximum to $w$. For any point on this path, the contour passing through this point must
be colored $c$.) Hence, $c \in \fcol(w)$ and $w \neq v$. Furthermore, $w$ is a critical point,
since $w$ is touched by another color. The corresponding edge cannot intersect $\phi$, so $w$
must have up-degree $2$.

All in all, when a new vertex is pushed onto the stack, it is always higher than the previous head.
\end{proof}

\medskip
\fbox{
\begin{minipage}{0.9\textwidth}
{\bf $\onestep(\MM)$}

\smallskip
\begin{compactenum}
	\item Run $\update(\st)$. Pop $\st$ to get $h$.
	\item Suppose $\mcol(h) = \{c\}$ (so $h$ is split). 
	\begin{compactenum}
		\item Connect $h$ (in $\reeb(\MM)$) to $\h(c)$ and the unique minimum corresponding to split $h$. 
		\item Delete $h$ from $T(c)$, set $\h(c) = h$.
		\item End procedure.
	\end{compactenum}
	\item Let $\mcol(h) = \{c_1, c_2\}$ (so $h$ is merge).
	\item Connect $h$ in $\reeb(\MM)$ to $\h(c_1)$ and $\h(c_2)$.
	\item Delete all copies of $h$ from $T(c_1)$ and $T(c_2)$.
	\item Perform union of colors $c_1$ and $c_2$ (denote merged color as $c$). Merge heaps
	to get $T(c) = T(c_1) \cup T(c_2)$.
	\item Set $\h(c) = h$.	
\end{compactenum}
\end{minipage}}

\pagebreak
\newcommand{\pathTree}{P_{\mathcal{S}}}
\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{S}}}
\section{Leaf assignments and path decompositions}

@@

>>>>

\newcommand{\build}{{\tt build}}
\newcommand{\col}{col}
\newcommand{\fcol}{full}
\newcommand{\cur}{cur}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\cost}{\mathop{cost}}
\newcommand{\h}{att}
\newcommand{\lift}{{\tt lift}}
\newcommand{\mcol}{mcol}
\newcommand{\merge}{{\tt merge}}
\newcommand{\onestep}{{\tt piece}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\pal}{P}
\newcommand{\pmax}{P_{\max}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{\cC}
\newcommand{\redH}{\widetilde{H}}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

We use key theorems from~\cite{Carr} (Theorem 6.6) that maps paths in $\cC(\MM)$ and $\MM$.

\begin{theorem} \label{thm:carr-path} For every path $P$ in $\MM$, there exists a path $Q$ in the contour tree corresponding
to the contours passing through points in $P$.
For every path $Q$ in the contour tree, there exists at least one path $P$ in $\MM$ through points present
in contours involving $Q$.
\end{theorem}

\begin{theorem} \label{thm:carr-mono} For every monotone path $P$ in $\MM$, there exists a monotone path $Q$ in the contour tree to which $P$
maps (as in the previous theorem), and vice versa.
\end{theorem}

@@

{\bf Painting:} The central tool is a notion 
of \emph{painting} of $\MM$. Initially associate a color with each maximum. Denoting the set of maxima is $X$,
this is an arbitrary bijection $\chi:X \mapsto [|X|]$. Imagine there being a large
can of paint of color $\chi(x)$ at maxima $x$. We will spill different paint from each maximum and watch it flow down.
This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
\emph{So paint only flows down edges, and it does not color the interior of facets.} Furthermore, paints
do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color. This process (and indeed the entire algorithm)
works purely on the 1-skeleton of $\MM$, which is just a graph.
%
\begin{definition} \label{def:paint} Let the 1-skeleton of $\MM$ be denoted $G(\MM) = (\cV,E)$.
A  \emph{painting} of $\MM$ is a map $\chi:X \cup E \mapsto [|X|]$ with the following property. 
 Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges along this path have the same color as $x$. (The maximum $x$
	is well-defined to be the unique maximum with the same color as $e$.)

An \emph{initial} painting has the additional restriction that the restriction $\chi:X \mapsto [|X|]$ is a bijection.
\end{definition}
%
Paintings are not unique. It is straightforward to construct an initial painting of $\MM$ in $O(|\MM|)$ time, 
using a descending BFS from each maximum that does not explore previously colored edges. 
As the algorithm proceeds, the bijection property will not hold, though we always maintain a valid painting.

@@

\begin{definition} \label{def:color-full} Consider vertex $v$. An \emph{up-star} of $v$ is the set 
of edges that all connected to a fixed component of $\MM^+_v$. (Each $v$
participates in at mot $2$ up-stars. A color $c \in \col(v)$ \emph{fully touches} $v$
if all edges in an up-star are colored $v$.
%and let $\phi$ be a $(f(v)+\delta)$-contour in
%$f|_{B_\eps(v)}$. If $\phi$ is monochromatic of color $c$, then $v$ is \emph{fully touched by $c$}. The set of colors
%that fully touch $v$ is denoted $\fcol(v)$.
\end{definition}

%Note that $\fcol(v)$ has size at most $2$, by the Morse condition. 
Note that $\col(v)$ 
%and $\fcol(v)$ can
be initially determined in $O(|\MM|)$ time, by a simple linear scan.

\begin{lemma} \label{lem:full} Suppose vertex $v$ is connected to a component of $\MM^+_v$
by edge $e$ of color $c$. Either all edges in $\MM^+_v$ are colored $c$, or there
exists critical vertex $w \in \MM^+_v$ fully touched by $c$ and touched by another color.
%
%and take $(f(v)+\delta)$-contour $\phi$ intersecting
%$B_\eps(v)$. Suppose the edges incident to $v$ intersecting $\phi$ do not have the same color. For any such
%color $c$, there exists a critical vertex $w$ strictly higher than $v$ such that $c \in \fcol(v)$.
\end{lemma}

\begin{proof} Since $e$ has color $c$,
there must exist vertices in $\MM^+_v$ touched by $c$. Consider the highest
vertex $w$ in $\MM^+_v$ that is touched by $c$ and some other color. If no such vertex exists,
this means all edges incident the a vertex touched by $c$ are colored $c$. By walking through
$\MM^+_v$, we deduce that all edges are colored $c$.

So assume $w$ exists. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
and intersect some $c$-colored edge incident to $w$. Note that all edges intersecting $\phi$ are also colored $c$,
since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
edges from the maximum to $w$. For any point on this path, the contour passing through this point must
be colored $c$.) Hence, $c$ fully touches $w$. 
But $w$ is touched by another color, and the corresponding edge cannot intersect $\phi$. So $w$
must have up-degree $2$ and is critical.
\end{proof}

\subsection{The overall approach and the data structures} \label{sec:struct}

We first list out the basic data structures used, and then summarize our approach.

\medskip
\noindent
{\bf The binomial heaps $\touch(c)$:} For each color $c$, $\touch(c)$ is a subset of vertices touched by $c$,
This is stored as a \emph{binomial max-heap} keyed by vertex heights. Abusing notation, $\touch(c)$ refers
both to the set and the data structure used to store it.

We initialize as follows. For each critical vertex $v$ and from each of its up-stars, we pick an arbitrary color $c$ touching $v$. 
We insert $v$ into $T(c)$. Observe that $v$ participates in at least $1$ and at most $2$ heaps.

@@

\medskip
\noindent
{\bf Attachment vertex $\h(c)$:} For each color $c$, we maintain a critical point $\h(c)$ of this color.
We will maintain the guarantee that the portion of the contour tree above (and including) $\h(c)$ has already been constructed.
This is initialized to the maximum corresponding to $c$.

\medskip
As the algorithm proceeds, it is convenient to imagine
it maintaining a set of contours $\Phi = \{\phi_1, \phi_2, \ldots, \phi_k\}$, which are contained
in incomparable edges of $\cC(\MM)$. Suppose we cut $\MM$ by each of these contours, and took
the simplicial complex $\NN$ where $\phi^-_i$'s are all boundary maxima. The algorithm will
have already determined the contour trees of the simplicial complices
where $\phi^+_i$ is a boundary maxima.

During this process, various colors will be \emph{merged}. This involves merging the respective $T(c)$'s and
updating the union-find data structure. The only operation performed on $T(c)$'s are delete-max and merge.
No new vertex is ever inserted into the heaps. As colors merge, the initial painting is obviously modified.
But it would be too expensive to explicitly update colors. So the new painting is given by $\rep(c)$,
while we only have access to the original painting. By union-find bounds, the only adds an $\alpha(|\MM|)$
overhead to the data structure operations. The new coloring is guaranteed to be a painting of $\NN$.

While this is the intuitive viewpoint, we describe how this information is actually maintained.
The vertices $\{\h(\rep(c))\}$ represent the set of contours $\Phi$. In terms of the contour tree, 
we have already constructed the contour trees rooted at each $h(\rep(c))$. This is trivially true at the initialization,
when each $h(c)$ is a maximum. One can now treat each $h(\rep(c))$ as a maximum, corresponding to (a possibly new) color $\rep(c)$.
To reduce confusion between the original painting and the new colors, we will assume that $\col(v)$ and $\fcol(v)$
are as given by the original painting.

\subsection{The algorithm} \label{sec:algo}

We formally describe the algorithm below. We require a technical definition of \emph{ripe} vertices.

\begin{definition} \label{def:ripe} A vertex $v$ is \emph{ripe} if: for all $c \in \col(v)$, $v$
is present in $T(\rep(c))$ and is also highest vertex in the heap. 
%Also, for all $c \in \col(v) \setminus \fcol(v)$, $T(\rep(c))$ is empty.
\end{definition}

\medskip
\fbox{
\begin{minipage}{0.9\textwidth}
{\bf $\build(\MM)$}

\smallskip
\begin{compactenum}
	\item While there are unprocessed critical points:
	\begin{compactenum}
		\item Run $\update(\st)$. Pop $\st$ to get $h$.
		\item Let $\cur(h) = \{\rep(c) | c \in \col(h)\}$.
		\item For all $c' \in \cur(h)$:
		\begin{compactenum}
			\item Add edge $(\h(c'),h)$ to $\cC(\MM)$.
			\item Delete $h$ from $T(c')$ (if present).
		\end{compactenum}
	\item Merge heaps $\{T(c') | c' \in \cur(h)\}$.
	\item Take union of $\cur(h)$ and denote resulting color by $\widehat{c}$.
	\item Set $\h(\widehat{c}) = h$ and mark $h$ as processed.
	\end{compactenum}
\end{compactenum}
\end{minipage}}

@@

\smallskip
\begin{compactenum}
	\item If $K$ is empty, push arbitrary unprocessed critical point $v$.
	\item Let $h$ be head of $\st$.
	\item While $h$ is not ripe:
	\begin{compactenum}
		\item Find $c \in \col(h)$ such that $h$ is not present or not the highest in $T(\rep(c))$.
		\item Push the highest of $T(\rep(c))$ onto $\st$, and update head $h$.
	\end{compactenum}
\end{compactenum}
\end{minipage}}

A few simple facts:
\begin{asparaitem}
	\item When $\update(\st)$ terminates, the head $h$ is ripe. 
	\item After $v$ is processed, all edges incident to $v$ has the same color.
\end{asparaitem}

\medskip
\ignore{	
	Suppose $\mcol(h) = \{c\}$ (so $h$ is split). 
	\begin{compactenum}
		\item Connect $h$ (in $\reeb(\MM)$) to $\h(c)$ and the unique minimum corresponding to split $h$. 
		\item Delete $h$ from $T(c)$, set $\h(c) = h$.
		\item End procedure.
	\end{compactenum}
	\item Let $\mcol(h) = \{c_1, c_2\}$ (so $h$ is merge).
	\item Connect $h$ in $\reeb(\MM)$ to $\h(c_1)$ and $\h(c_2)$.
	\item Delete all copies of $h$ from $T(c_1)$ and $T(c_2)$.
	\item Perform union of colors $c_1$ and $c_2$ (denote merged color as $c$). Merge heaps
	to get $T(c) = T(c_1) \cup T(c_2)$.
	\item Set $\h(c) = h$.

We state the primary invariant below.
First, some notation. For any $v$, let $\psi^-_v$ denote the $(f(v)-\delta)$-contour intersecting
$B_\eps(v)$. If there are two such contours (so $v$ is a split), choose the one that contains
the dominant minimum. The \emph{palette} $\pal$ is the set of colors currently used,
which is $\{T(\rep(c)) | c \in |X|\}$.

Observe that $\build(\MM)$ loops over all unprocessed critical point.
The invariant is true at the starting point of each such iteration.

\medskip
\textbf{Invariant:} 
\begin{compactenum}
	\item For every $c \in \pal$, the subtree of $\cJ(\MM)$ rooted at $\h(c)$ has been constructed.
	\item Fix $c \in \pal$. Consider the set $S$ of maxima of $\MM$ in the subtree of $\cJ(\MM)$ rooted at $\h(c)$.
	Then $\rep(c) = \bigcup_{s \in S} \chi(s)$.
	\item Let $\Psi = \{\psi^-_{\h(c)} | c \in \pal\}$. The coloring given by $\rep(\cdot)$
	is a valid painting of $\cut(\MM,\Psi)$, where $\psi^-_{\h(c)}$ has color $c$.
\end{compactenum}

\medskip
It is easy to see that the invariant is true at the very beginning of the algorithm.
Each $\h(c)$ is simply the maximum colored with $c$, and we have a valid painting of $\MM$.
}

\section{Proving correctness} \label{sec:correct}

\ignore{
\begin{claim} \label{clm:process} Assume the invariant. Any vertex with a non-increasing path to some $\h(c)$
for $c \in \pal$ has been processed.
\end{claim}

\begin{proof} This is a direct consequence of \Thm{carr-mono}, which relates monotone paths in $\MM$ to $\cC(\MM)$.
Since the subtree of $\cJ(\MM)$ (which is basically the subtree of $\cC(\MM)$) rooted at $\h(c)$ 
has been found, all vertices in this subtree must be processed. These are all the vertices with non-increasing paths
to $\h(c)$ in $\cC(\MM)$, which by \Thm{carr-mono} is the same as those in $\cC(\MM)$.
\end{proof}
}

The following crucial claim demonstrates the utility of the stack.

@@

\begin{proof} Consider a single push onto the stack and the current painting. We have $c \in \col(h)$, as given in pseudocode of $\update$.
If $h \in T(\rep(c))$, then obviously the next head is higher. So assume that $h \notin T(\rep(c))$.
There is obviously an edge of color $\rep(c)$ touching $h$ connected to a component $\SS$ of $\MM^+_v$. 
From the initialization, $h \in T(\rep(c'))$, for some $c'$ touching $h$ (where $c'$ is the color
of some edge connecting $h$ to $\SS$).
Applying \Lem{full},  either all edges in $\SS$ are colored $\rep(c)$ or 
there exists critical vertex $w \in \SS$ fully touched by $\rep(c)$ and another color.
In the former case, all edges connecting $\SS$ to $h$ are also colored $\rep(c)$
(since this is a painting). Hence, $\rep(c') = \rep(c)$ and $h \in T(\rep(c))$.
As we assumed this is not the case, we conclude the existence of critical vertex $w$.
Since $w$ is fully touched by $\rep(c)$, $w \in T(\rep(c))$. Since $w$
is touched by another color, $w$ has not been processed. So the highest of $T(\rep(c))$
is higher than $h$, as desired.
\ignore{
If all edges touching $h$ that connect to this component are also colored $\rep(c)$, then
In Case (i), the increase in height is obvious.
Consider Case (ii). There is a vertex $v$ touched by color $c \notin \fcol(v)$. Consider the highest
vertex $w$ that is touched by $c$ and some other color. Clearly such a vertex exists, since the condition
is also satisfied by $v$. Take the $(f(w)+\delta)$-contour $\phi$ that intersects $B_\eps(v)$
and intersect some $c$-colored edge incident to $w$. Note that $\phi$ is also colored $c$,
since $w$ is the highest vertex to be touched by $c$ and some other color. (Take the path of $c$-colored
edges from the maximum to $w$. For any point on this path, the contour passing through this point must
be colored $c$.) Hence, $c \in \fcol(w)$ and $w \neq v$. Furthermore, $w$ is a critical point,
since $w$ is touched by another color. The corresponding edge cannot intersect $\phi$, so $w$
must have up-degree $2$.
All in all, when a new vertex is pushed onto the stack, it is always higher than the previous head.
}
\end{proof}

\begin{claim} \label{clm:ripe} For ripe vertex $v$, $\{\h(\rep(c)) | c \in \col(v)\}$ is exactly
the set of parents of $v$.
\end{claim}

\begin{proof} \Sesh{proof is a little sketchy.}
Suppose $v$ is ripe. For any $c$ touching $v$, $v$ is the highest in $T(\rep(c))$.
Consider some such $\rep(c)$, for an edge connecting to component $\SS$ of $\MM^+_v$.
Either all edges in $\SS$ are colored $\rep(c)$ or there exists critical vertex $w \in \SS$
fully touched by $\rep(c)$ and another color. In the latter case, $w$ has not been processed,
so $w \in T(\rep(c))$ (contradiction). Therefore, all edges in $\SS$ are colored $\rep(c)$.
Suppose there is some critical vertex in $\SS$. If it is not processed yet, it will
be in $T(\rep(c))$, which is not possible. Thus, all critical vertices in $\SS$ have been processed.

Consider $w$, the lowest critical vertex in $\SS$. When $w$ was processed, it was set to $\h(c')$, for
some color $c'$. At this stage, all edges in $\SS$ above $w$ were also colored $c'$ (just apply the argument
above for $w$). Since there are no critical points below $w$, all edges in $\SS$ below $w$
are also colored $c'$ (the painting property). This means that the edges connecting $\SS$
are also colored $c'$. So $c' = \rep(c)$, and $\h(\rep(c)) = w$ is a parent of $v$.

This argument can be repeated for all such components $\SS$ to complete the proof.
\end{proof}

\newcommand{\pathTree}{P_{\mathcal{S}}}
\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{S}}}
\section{Leaf assignments and path decompositions}

@@

\section{Old proof stuff}

\begin{claim}
Suppose ripe vertex $v$ is connected to a component of $\MM^+_v$. For every edge $e$ incident to a vertex of $\MM^+_v$,
$\rep(\chi(e))$ is identical. Furthermore, all vertices in $\MM^+_v$ have been processed.
\end{claim}

\begin{proof} Consider the painting of the lower simplicial complex $\cut(\MM,\Psi)$ (call it $\NN$). 
By the invariant, $\rep(\cdot)$ provides a valid coloring. Note that $v$ is not processed (it is ripe),
so if we take $(f(v)+\delta)$-contours intersecting $B_\eps(v)$, these contours could not have been
processed. Therefore, there must exist a non-trivial component of $\NN^+_v$ that $v$
is connected to. By \Lem{full}, either all edges in this component have the same color
or

Suppose ripe vertex $v$
is connected to some component in $\NN^+_v$. 
Consider edge $e$ incident to $v$ intersecting $\phi$ of color $c$. 
By \Lem{full}, there exists critical vertex $w \in \MM^+_v$ such that $w \in \fcol(c)$. 
Regardless of whether $v \in \fcol(c)$ or not, the vertex $w$ must have been processed.

\pagebreak
